---
title: "Assignment 10: Data Scraping"
author: "Jonathan Joyner"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)
library(knitr)
getwd()

# Set theme
mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
DurhamLWSP <- read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022")
DurhamLWSP
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
Water_System_Name <- DurhamLWSP %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
Water_System_Name

DurhamPWSID <- DurhamLWSP %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
DurhamPWSID

Ownership <- DurhamLWSP %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
Ownership

MaxDayUse <- DurhamLWSP %>%
  html_nodes("th~ td+ td") %>% 
  html_text()
MaxDayUse
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 AI assisted for date object transformation code
custom_month_order <- c("JAN","MAY","SEP",
                        "FEB","JUN","OCT",
                        "MAR","JUL","NOV",
                        "APR","AUG","DEC")

Durham_df <- data.frame("Water_System_Name" = as.character(Water_System_Name),
                        "PWSID" = as.character(DurhamPWSID),
                        "Ownership" = as.character(Ownership),
                        "Max_Day_Use" = as.numeric(MaxDayUse),
                        "Month" = custom_month_order,
                        "Year" = rep(2022))
Durham_df$Date <- 
  as.Date(paste(Durham_df$Year, Durham_df$Month, "01", sep = "-"),
                format = "%Y-%b-%d")
view(Durham_df)
kable(Durham_df)

#5 AI assisted
ggplot(Durham_df, aes(x = Date, y = Max_Day_Use)) +
  geom_line() +
  labs(title = "Maximum Daily Withdrawals (2022)",
       x = "Date",
       y = "Max Daily Withdrawals") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month")
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
scrape.it <- function(the_year,the_facility){

the_website <- read_html(paste0(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=",
  the_facility,"&year=",the_year))
  
the_facility_name_tag <- "div+ table tr:nth-child(1) td:nth-child(2)"
the_PWSID_tag <- "td tr:nth-child(1) td:nth-child(5)"
the_Max_Day_Use_tag <- "th~ td+ td"
the_Ownership_tag <- "div+ table tr:nth-child(2) td:nth-child(4)"

the_facility_name <- the_website %>% html_nodes(the_facility_name_tag) %>% html_text()
the_PWSID <- the_website %>% html_nodes(the_PWSID_tag) %>% html_text()
the_Max_Day_Use <- the_website %>% html_nodes(the_Max_Day_Use_tag) %>% html_text()
the_Ownership <- the_website %>% html_nodes(the_Ownership_tag) %>% html_text()

NCDIV_water_scrape <- data.frame(
                        "Water_System_Name" = as.character(the_facility_name),
                        "PWSID" = as.character(the_PWSID),
                        "Ownership" = as.character(the_Ownership),
                        "Max_Day_Use" = as.numeric(the_Max_Day_Use),
                        "Month" = custom_month_order,
                        "Year" = as.numeric(the_year))
NCDIV_water_scrape$Date <- 
  as.Date(paste(NCDIV_water_scrape$Year, NCDIV_water_scrape$Month, "01", sep = "-"),
                format = "%Y-%b-%d")
return(NCDIV_water_scrape)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
Durham_2015 <- scrape.it(2015,"03-32-010")
view(Durham_2015)
kable(Durham_2015)

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 AI Assisted
Asheville_2015 <- scrape.it(2015,"01-11-010")
view(Asheville_2015)
kable(Asheville_2015)

combined_data <- rbind(Durham_2015, Asheville_2015)

ggplot(combined_data, aes(x = Date, y = Max_Day_Use, color = Water_System_Name)) +
  geom_line() +
  labs(title = "Maximum Daily Withdrawals (2015)",
       x = "Date",
       y = "Max Daily Withdrawals") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 AI Assisted
Ash_years <- 2010:2021
Ash_facility_id <- "01-11-010"

combined_data_asheville <- 
  map2(Ash_years, rep(Ash_facility_id, length(Ash_years)), scrape.it) %>%
  bind_rows()

combined_data_asheville$Year <- as.factor(combined_data_asheville$Year)

ggplot(combined_data_asheville, aes(x = Date, y = Max_Day_Use)) +
  geom_line(aes(color = Year), size = 1, alpha = 0.7) +
  geom_point(aes(color = Year), size = 1, shape = 19,color = 'black') +
  geom_smooth(method = 'loess', se = FALSE, color = 'blue', size = 1) +
  labs(title = "Asheville's Maximum Daily Withdrawals (2010-2021)",
       x = "Date",
       y = "Max Daily Withdrawals") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  theme(panel.grid.major.y = element_line(color = "gray", linetype = "dashed"))

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer: Ashville was on a downward trend until 2015, but then started trending up. In 2019, water withdrawl spiked noticeably.
