---
title: "Assignment 5: Data Visualization"
author: "Jonathan Joyner"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
#import packages
library(tidyverse)
library(lubridate)
library(cowplot)
library(ggplot2)
#set path for processed data files 
getwd()
setwd(
  "C:/Users/jbjoy/OneDrive/Documents/Grad School/Fall 2023/Environ 872/EDE_Fall2023/Data/Processed_KEY")
getwd()
#read processed data files
PeterPaul<-
  read.csv(
    "NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv",
    stringsAsFactors = TRUE)
NIWOT<-
  read.csv(
    "NEON_NIWO_Litter_mass_trap_Processed.csv",
    stringsAsFactors = TRUE)
#2 
#make sure date columns are read as date factor
PeterPaul$sampledate<-ymd(PeterPaul$sampledate)
NIWOT$collectDate<-ymd(NIWOT$collectDate)
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme and plot total P vs PO4}
#3 adjust theme 
d_theme<-theme_classic(base_size = 14) +
  theme(axis.text= element_text(color="blue"),
        legend.position="top",
        legend.justification ="center")
#set theme
theme_set(d_theme)

#4 Plot total by phosphate and set aesthetics with best fit line
#set vertical limit of 50
Plot4<-
  ggplot(PeterPaul,aes(x = tp_ug, y=po4,color=lakename))+
  geom_point()+
  geom_smooth(method=lm,color="black") +
  ylim(0,50)
print(Plot4)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 Rename axis labels
#create boxplot by temperature
Plot5a<-
  ggplot(PeterPaul,
         aes(x=factor(month,levels=1:12,labels=month.abb),
             y = temperature_C,
             color=lakename))+
  geom_boxplot() +
  labs(y="Tempereature",x="") +
  theme(legend.position ="bottom")
print(Plot5a)
#create boxplot by TP
Plot5b<-
  ggplot(PeterPaul,
         aes(x = factor(month,levels=1:12,labels=month.abb),
             y = temperature_C,
             color=lakename))+
  geom_boxplot() +
  labs(y="TP",x="") +
  theme(legend.position ="bottom")
print(Plot5b)
#create boxplot by TN
Plot5c<-
  ggplot(PeterPaul,
         aes(x = factor(month,levels=1:12,labels=month.abb),
             y = temperature_C,
             color=lakename))+
  geom_boxplot() +
  labs(y="TN",x="") +
  theme(legend.position ="bottom")
print(Plot5c)
#Combine 3 boxplots, removing legends and adding back modified legend
Plot5<-
  plot_grid(Plot5a + theme(legend.position = "none"),
            Plot5b + theme(legend.position = "none"),
            Plot5c + theme(legend.position = "none"),
  get_legend(Plot5a),
  ncol=1,
  align="vh",
  axis="b",
  rel_heights=c(1,1,1,0.5))
print(Plot5)
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: The levels of of TP and TN correlate strongly with seasonal temperature, rising in the summer and falling in winter.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
#Filter NIWOT data for needles
NIWOTNeedles<-
  NIWOT%>%
  filter(functionalGroup=="Needles")
#Plot filtered data by date, mass, and class
Plot6F<-
  ggplot(NIWOTNeedles,
         aes(x = collectDate,
             y = dryMass,
             color=nlcdClass))+
  geom_point()
print(Plot6F)
#7
Plot7 <-
  ggplot(NIWOTNeedles,
         aes(x = collectDate,
             y = dryMass)) +
  geom_point() +
  facet_wrap(vars(nlcdClass))
print(Plot7)
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I believe 7 is more effective because viewing the data by isolated class is easier to understand than trying to color code data. I this case, the color coded data in 6 ends up being both crowded and obscured.