---
title: "Assignment 5: Data Visualization"
author: "Jonathan Joyner"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
#import packages
library(tidyverse)
library(lubridate)
library(cowplot)
library(ggplot2)
#set path for processed data files
setwd("C:/Users/jbjoy/OneDrive/Documents/Grad School/Fall 2023/Environ 872/EDE_Fall2023/Data/Processed_KEY")
#read processed data files
PeterPaul<-
  read.csv("NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv",stringsAsFactors = TRUE)
NIWOT<-
  read.csv("NEON_NIWO_Litter_mass_trap_Processed.csv",stringsAsFactors = TRUE)
#2 
#make sure date columns are read as date factor
PeterPaul$sampledate<-ymd(PeterPaul$sampledate)
NIWOT$collectDate<-ymd(NIWOT$collectDate)
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
d_theme<-theme_classic(base_size = 14) +
  theme(axis.text= element_text(color="blue"),
        legend.position="top",
        legend.justification ="center")
theme_set(d_theme)

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
Plot4<-
  ggplot(PeterPaul,aes(x = tp_ug, y=po4,color=lakename))+
  geom_point()+
  geom_smooth(color="black") +
  ylim(0,50)
print(Plot4)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 
#create boxplot by temperature
Plot5a<- ggplot(PeterPaul,aes(x = month, y = temperature_C, color=lakename))+
  geom_boxplot()
print(Plot5a)
#create boxplot by TP
Plot5b<- ggplot(PeterPaul,aes(x = month, y = tp_ug, color=lakename))+
  geom_boxplot()
print(Plot5b)
#create boxplot by TN
Plot5c<- ggplot(PeterPaul,aes(x = month, y = tn_ug, color=lakename))+
  geom_boxplot()
print(Plot5c)
#Isolate legend
Master_Legend<-get_legend(Plot5a)
#Combine 3 boxplots
Plot5<-
  plot_grid(
    Plot5a+theme(legend.position = "none"),
    Plot5b+theme(legend.position = "none"),
    Plot5c+theme(legend.position = "none"),
    nrow = 1, align = "h") +
  theme(Master_Legend)
print(Plot5)
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
NIWOTNeedles<-
  NIWOT%>%
  filter(functionalGroup=="Needles")
Plot6F<- ggplot(NIWOTNeedles,aes(x = collectDate, y = dryMass, color=nlcdClass))+
  geom_point()
print(Plot6F)
#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: